Mountaineering: Be lucky first, prepared second

### 2024-12-04 paper/paper.qmd
Save paper, code style changes to model data

The Himalayan mountain range spans X KM (source), which means that the graphed distribution does not represent the distribution of ALL peaks in the Himalayas. Instead this distribution represents the distribution of peaks commonly attempted/climbed.The Himalayas starting at X M starting above sea level combined with the lowest height required to be considered a peak, the distribution starts at 5000m above sea level.

@@

peaks |>
  ggplot(aes(climbing_status, fill = climbing_status)) +
  geom_bar() +
  labs(
    title = "More than a quarter of Himalayan peaks remain unclimbed",
    x = "",
    y = "Number of peaks"
  ) +
  theme(legend.position = "none")
```
This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.

\newpage
Why do so many climbs fail however? The following graph shows the most common reasons.

```{r,fig.width = 12, fig.height = 9}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-termination
#| tbl-cap: "Top 5 most frequent pollsters, with count of polls, average pollscore (lower scores indicate less bias), and average numeric grade (higher values indicate greater reliability)."

@@

@tbl-termination shows that most attempts fail due to partially random events like weather, or random events such Avalanches. 
\newpage

```{r, fig.width = 12, fig.height = 3}
#| warning: false
#| message: false
#| echo: false
#| label: fig-summit
#| fig-cap: "Distribution of First ascents across years"

@@

@fig-summit shows that while there was an intial surge then drop off of first ascents, there was again a steady increase in first ascents actually peaking in 2019, and the 2020 data was inconclusive since not all of the year was recorded in the data. This shows that first ascents are currently trending upwards
\newpage

```{r,fig.width = 12, fig.height = 6}
#| warning: false
#| message: false
#| echo: false
#| label: fig-trend
#| fig-cap: "Distribution of first ascents by country of citizenship"

@@

@fig-trend Expectly most first ascents are from the country that the Himalayas reside in, followed by Japan, UK, France, Germany, then the USA..
\newpage

@@

## Measurement

@@

## Model Justification

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary1

@@

modelsummary(model2,statistic = c("p.value"))
```
\newpage
\newpage

shows the influence (top number) and P-Value (bottom number in brackets) of each variable. For example each year increase in age reduces the probability they will succeed by around two percent with a P-value of 0.001 which means we are very sure of this relationship. A P-value of 0.001 means only 0.1% of the time this relationship can be explained by random chance. Because the standard level of uncertainty is less than 5% we should try to remove variables over a 0.05 P-value. There are also several figures relating to model accurary at the bottom of the figure;the main one to focus on is RMSE. An RMSE of 0.41 means the model predicts results within 41% of the actual results. For a model with a binary outcome it means our model predicts successes 59% of the time.]

@@

modelsummary(model1, statistic = c("p.value"))
```
\newpage
\newpage

shows us that despite removing citizenship and hired_staff the RMSE only grains 0.01, meaing it loses 1% accuracy. This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.
```{r, fig.width = 2, fig.height = 1}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary

@@

# Appendix {#sec-appendix}
## Cleaning
For the analysis data, the cleaning steps I took were:

@@

>>>>

The Himalayan mountain range spans a large amount of distance, which means that the graphed distribution does not represent the distribution of ALL peaks in the Himalayas. Instead this distribution represents the distribution of peaks commonly attempted/climbed.The Himalayas starting at X M starting above 2000m sea level combined with the lowest height required to be considered a peak, the distribution starts at 5000m above sea level.

@@

peaks |>
  ggplot(aes(climbing_status, fill = climbing_status)) +
  geom_bar() +
  labs(
    title = "More than a quarter of Himalayan peaks remain unclimbed",
    x = "",
    y = "Number of peaks"
  ) +
  theme(legend.position = "none")
```
This graph shows that despite over 100 years of climbing history, not all attempted peaks have been conquered.
Why do so many climbs fail however? The following graph shows the most common reasons.
\newpage

```{r,fig.width = 12, fig.height = 9}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-termination
#| tbl-cap: "Common reasons why attempts fail"

@@

@tbl-termination shows that most attempts fail due to partially random events like weather, or random events such Avalanches. 
\newpage
```{r, fig.width = 12, fig.height = 3}
#| warning: false
#| message: false
#| echo: false
#| label: fig-summit
#| fig-cap: "Distribution of First ascents across years"

@@

@fig-summit shows that while there was an intial surge then drop off of first ascents, there was again a steady increase in first ascents actually peaking in 2019, and the 2020 data was inconclusive since not all of the year was recorded in the data. This shows that first ascents are currently trending upwards

```{r,fig.width = 12, fig.height = 5}
#| warning: false
#| message: false
#| echo: false
#| label: fig-trend
#| fig-cap: "Distribution of first ascents by country of citizenship"

@@

@fig-trend Expectedly most first ascents are from the country that the Himalayas reside in, followed by Japan, UK, France, Germany, then the USA..

@@

## Measurement
The measurement of Himalayan climbing data in the referenced dataset involves converting real-world phenomena—such as expeditions, summits, and fatalities—into structured entries. This process begins with records collected from expedition reports, which are typically submitted by climbers or organizations. These reports include details about expedition dates, routes, climbers' nationalities, success in reaching the summit, and any fatalities or injuries that occurred.

The dataset relies heavily on sources like The Himalayan Database, a comprehensive archive maintained by Elizabeth Hawley and her successors. This database aggregates information from direct interviews with climbers, expedition documentation, and third-party reports. However, the translation of these real-world events into dataset entries introduces several challenges. Discrepancies may arise due to incomplete or inconsistent reporting, varying definitions of a “summit,” and the underrepresentation of smaller, independent expeditions.

For example, an expedition is recorded in the dataset only if sufficient documentation is available to verify its occurrence and outcomes. Factors such as weather conditions, local incidents, or logistical difficulties that don't result in a summit or a fatality may not be consistently recorded. Additionally, subjective elements, like the climbers’ or organizers' interpretation of success, influence the data's structure. These steps, while thorough, highlight potential biases in how phenomena are represented as measurable data.

@@

## Model Justification
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary1

@@

#modelsummary(model2,statistic = c("p.value")) simply took output and inputted as png, the length of the table is too long and won't wrap
```
\newpage

![](image1.png)
This table shows the influence (top number) and P-Value (bottom number in brackets) of each variable. For example each year increase in age reduces the probability they will succeed by around two percent with a P-value of 0.001 which means we are very sure of this relationship.
A P-value of 0.001 means only 0.1% of the time this relationship can be explained by random chance. Because the standard level of uncertainty is less than 5% we should try to remove variables over a 0.05 P-value. There are also several figures relating to model accurary at the bottom of the figure;the main one to focus on is RMSE. An RMSE of 0.41 means the model predicts results within 41% of the actual results. For a model with a binary outcome it means our model predicts successes 59% of the time.(removed entries for each country due to formatting issues)

@@

modelsummary(model1, statistic = c("p.value"))
```
\newpage
The second table shows the results when citizenship and hired_staff are removed. We see the RMSE only grains 0.01, meaning it loses 1% accuracy. This means the citizenship of climber nor the amount of hired staff did not contribute to the model's accuracy and were good to remove.
```{r, fig.width = 2, fig.height = 1}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary

@@

# Appendix {#sec-appendix}
## Cleaning
For the analysis data, the cleaning steps were from @Cookson

@@

## Observational data
Himalayan climbing data, particularly concerning summits and fatalities on peaks like Mount Everest, may misrepresent actual outcomes due to several factors, including incomplete reporting, commercial biases, and evolving definitions of success. These discrepancies can create a skewed understanding of the risks and achievements in high-altitude mountaineering.
Incomplete Reporting

Many summits, fatalities, and near-misses in the Himalayas go unreported or inaccurately documented due to the remote and challenging environments of the region. Smaller expeditions, particularly those from less prominent countries or by independent climbers, may not submit detailed records. Additionally, incidents on less-frequented peaks are less likely to be covered in international or even regional climbing databases, leading to underrepresentation in global statistics. For example, organizations like the Himalayan Database, while comprehensive, rely heavily on self-reported data, which can vary in reliability depending on the source.
Commercial Influences

The rise of commercial expedition companies has significantly influenced how climbing data is presented. For marketing purposes, these companies may highlight success rates while downplaying or omitting fatalities or unsuccessful attempts. Moreover, certain deaths that occur before or after the official climbing season might not be included in annual tallies, further skewing statistics. High-profile peaks like Everest often see overrepresentation in the data, as their fame and accessibility make them more frequently reported than other Himalayan mountains.

The definition of a “summit” itself can vary, adding further complications. On Everest, for example, some climbers might claim summits upon reaching the South Summit or other points close to the highest peak, particularly in cases of bad weather or emergencies. Such claims are sometimes accepted without verification, leading to inflated success statistics. Disputes over what constitutes a legitimate ascent, particularly as aided by modern technology like oxygen tanks and fixed ropes, also muddy the waters of accurate representation.
Bias in Reporting Fatalities

Fatality rates may also be misrepresented due to a focus on high-profile accidents involving Western climbers or commercial expeditions, while incidents involving Sherpas and local climbers are often underreported. Sherpas, who take on the riskiest roles during climbs, such as route preparation and heavy load carrying, face disproportionate risks that may not be fully reflected in the data.
Conclusion

The misrepresentation of Himalayan climbing data has implications for understanding the true risks and challenges of high-altitude mountaineering. Efforts to improve the accuracy of this data, such as independent verification and more comprehensive reporting systems, are essential for providing climbers, researchers, and policymakers with reliable information. Sources like The Himalayan Database [@Salisbury], the American Alpine Journal, and recent academic studies on mountaineering ethics and statistics provide valuable insights but also highlight the need for more transparent and standardized data collection practices.
